\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{url}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Prompt-Conditioned Naturalness Distributions for Speech-to-Speech Model Evaluation\\
\thanks{This work was supported by the SPEAR project. Add full funding acknowledgments before submission.}
}

% \author{Anonymized submission to IEEE SLT 2026}
\author{\IEEEauthorblockN{Ashish Hallur\IEEEauthorrefmark{1}, Sathvik Manikantan Napa Ugandhar\IEEEauthorrefmark{1}, Thomas Thebaud\IEEEauthorrefmark{1}, \\ 
Georgi Tinchev\IEEEauthorrefmark{2}, Venkatesh Ravichandran\IEEEauthorrefmark{2}, Laureano Moro-Velazquez\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\IEEEauthorrefmark{1}\textit{Department of Electrical and Computer Engineering, Johns Hopkins University}, Baltimore, MD, USA\\
Email: ahallur1@jhu.edu}
\IEEEauthorblockA{\IEEEauthorrefmark{2}\textit{Amazon AGI}, Seattle, WA, USA}}

\maketitle

\begin{abstract}
Speech-to-speech (S2S) systems are increasingly capable of producing fluent spoken responses, but their evaluation remains weakly tied to the distributional structure of natural conversation.
We propose a prompt-conditioned evaluation framework that learns the expected distribution of response-side speech metrics from natural dyadic interactions.
Given current prompt-side acoustic, temporal, affective, speaker-profile, conversational-context, speech-embedding, and multiturn history features, the model predicts a distribution over target response metrics including duration, lexical length, prosody, rhythm, pausing, affect, and speaker-profile estimates.
An S2S output can then be scored by extracting the same response metrics and comparing them against the learned conditional distribution for the corresponding prompt.
We report a multiturn conditional variational autoencoder (CVAE) sweep over history windows of one, three, five, and seven previous turns, comparing tabular-only conditioning, Trillsson embeddings, Whisper embeddings, and dual embedding fusion.
The best validation result is obtained by dual embedding fusion with a three-turn history window, which achieves the lowest scaled mean squared error (MSE) and highest $R^2$ among the evaluated variants.
We also define a prompt-response human evaluation protocol to validate whether automatic naturalness scores align with listener judgments of response naturalness, conversational fit, timing, prosody, and signal quality.
\end{abstract}

\begin{IEEEkeywords}
speech-to-speech evaluation, conversational naturalness, conditional generative modeling, speech embeddings, variational autoencoders
\end{IEEEkeywords}

\section{Introduction}
Spoken dialogue is shaped by more than lexical content.
Natural responses occupy context-dependent ranges of duration, speaking rate, pausing, pitch behavior, affective tone, and speaker-state cues.
These dimensions are central to turn-taking and interactional coordination, where speakers use timing, prosody, and syntactic structure to manage smooth exchanges \cite{riest_anticipation_2015, levinson_turn-taking_2016, koiso_analysis_1998}.
As S2S systems move from read-speech synthesis toward interactive spoken agents, evaluation methods must capture whether an output is plausible for the local conversational context rather than merely intelligible or textually correct \cite{omahony_combining_2022}.

Many current evaluation practices for spoken dialogue systems remain centered on automatic speech recognition, task completion, text similarity, or global human ratings.
These signals are useful, but they do not directly explain why a generated response sounds natural or unnatural as speech.
For example, an S2S response may be semantically appropriate but too slow for a high-arousal prompt, too monotone for an expressive exchange, or temporally mismatched to the local turn transition.
Such failures are naturally described as deviations from conditional distributions over speech behavior.

We construct these reference distributions directly from the Seamless Interaction dataset by converting adjacent cross-speaker turns into prompt-response examples and extracting the same set of metrics from both sides of each pair.
The resulting tabular training data combines transcript-derived timing, VAD-refined segment boundaries, Praat-based pitch measurements, local pause and rate measures, WavLM-based speaker-profile and affect estimates, dyadic context, prompt speech embeddings, and previous-turn history.
Rather than defining one pooled acceptable range for each metric, we learn a history-conditioned distribution over response-side metrics.
At evaluation time, an S2S waveform is mapped into the same metric space and scored by its likelihood, calibrated interval membership, and per-feature standardized residuals under the predicted distribution.

The central object is a conditional model
\begin{equation}
    p_{\theta}(y_t \mid x_t, h_t^{(H)}),
    \label{eq:conditional_distribution}
\end{equation}
where $x_t$ contains the current prompt-side acoustic, temporal, affective, speaker-profile, context, and embedding features, $h_t^{(H)}$ is a padded sequence of up to $H$ previous turns, and $y_t$ contains response-side target metrics for the current turn.
The model conditions only on the current prompt and available interaction history.
The S2S response itself is not used as conditioning information; it is the object being scored.

This paper contributes a distributional formulation for S2S naturalness evaluation, a paired prompt-response metric pipeline over natural conversational speech, and a multiturn CVAE framework that compares tabular-only, Trillsson, VoxProfile/Whisper, and dual-embedding conditioning.
We further specify a prompt-conditioned human evaluation protocol that compares natural human-human responses, deliberately unnatural EmotiVoice responses, and neutral human-S2S agent responses under the same listener-rating interface.
Across a 16-run validation sweep, dual embedding with $H=3$ produces the best MSE and $R^2$, reducing MSE by 3.90\% relative to the strongest tabular-only multiturn baseline.

\section{Related Work}
\subsection{Speech-Native S2S Evaluation}
S2S evaluation has often inherited methods from automatic speech recognition, text generation, and text-to-speech evaluation.
These approaches measure important aspects of system quality, but they can underrepresent conversational timing, turn-taking fit, prosodic expressivity, and context-appropriate response behavior.
Subjective ratings remain essential, yet they are expensive to collect and often provide limited diagnostic detail.
Recent work on conversational agents and speech synthesis has further shown that speech rate, prosody, and interaction partner can shape user experience and spoken behavior, reinforcing the need for evaluation signals that remain in the speech domain \cite{dowding_user_2024, xie_speaking_2024, jung_comparison_2025}.
The goal of the present work is not to replace human evaluation, but to provide a scalable speech-native diagnostic that can identify which measurable dimensions of a response deviate from natural conversational behavior.

\subsection{Conversational Prosody and Rhythm}
Natural conversation is coordinated through timing and prosodic cues that help manage turns, signal stance, and shape perceived engagement.
Prior work has shown that conversational speech differs from read speech in both prosodic and temporal structure, and that speaking rate, pausing, and pitch variation shift with speaker traits and interactional state \cite{howell1991comparison, hazan2010does, tauroza_speech_1990, goldman1958speech, wynn_conversational_2024}.
Prosodic variation, particularly $F_0$ level and range, is associated with communicative intent, emotional expression, stress, and discourse structure \cite{banziger_role_2005, pell_emotional_2011, xu_speech_2005, sluijter_acoustic_1996, sluijter_spectral_1996}.
Temporal rhythm similarly depends on pause structure and speaking-rate variation, which have been studied in spontaneous and conversational speech across multiple measurement settings \cite{campione2002large, pellegrino_automatic_2004, jiao_estimating_2015, morgan_combining_1998}.
This motivates an evaluation design in which naturalness is judged relative to a matched prompt and context, not against a single global baseline.
In this work, the empirical reference space is built within the current pipeline: conversational turns are segmented, filtered, aligned to word timing and VAD, converted into adjacent prompt-response pairs, and represented with interpretable prosodic, rhythmic, affective, and profile-derived measurements.

\subsection{Distributional Modeling}
Point predictions are insufficient when the target object is a range of plausible responses.
For S2S evaluation, multiple different response patterns may be natural for the same prompt, and a model should represent this uncertainty.
We therefore use probabilistic models that produce conditional distributions.
CVAE models use latent variables to represent response variability conditioned on prompt context and history \cite{kingma2013vae, sohn2015cvae}.

\subsection{Speech Embeddings}
Hand-engineered acoustic and temporal metrics summarize interpretable dimensions of speech, but they may miss broader information about prompt content, speaking style, and paralinguistic context.
We therefore incorporate learned prompt representations from two complementary embedding families.
VoxProfile/Whisper embeddings provide 1,280-dimensional representations derived from hidden states of a Whisper-based speech model \cite{radford2023whisper}.
Trillsson embeddings provide 1,024-dimensional non-semantic speech representations designed for paralinguistic information \cite{shor2022trillsson}.
In the proposed evaluator, these embeddings augment the prompt-side conditioning vector rather than replacing interpretable metric features.

\section{Dataset and Metrics}
\subsection{Paired Prompt-Response Construction}
We construct paired examples from the Seamless Interaction dataset, a large corpus of dyadic English conversations containing naturalistic and improvised interactions.
The corpus contains thousands of hours of face-to-face dyadic interaction, making it suitable for estimating conversational reference distributions at scale \cite{seamless_interaction}.
Each training example corresponds to an adjacent prompt-response pair.
The prompt segment provides current-turn conditioning information, and the following response segment provides the target metric vector.

The pair construction pipeline begins from word-level transcript events and dyad metadata.
Transcript events from the same speaker are merged when separated by at most 1.0 s, and VAD regions are merged using the same gap threshold.
Transcript-derived segment boundaries are then refined against VAD regions that overlap the event or fall within a 0.5 s tolerance.
Candidate prompt-response examples are formed from adjacent segments only when the speaker changes.
Very short backchannel turns are removed using a fixed list of common acknowledgments, and pairs are filtered to keep segments between 3 and 15 s.
If a candidate prompt or response is longer than 15 s, the prompt keeps the final 15 s before the turn transition and the response keeps the first 15 s after the transition.
Pairs with more than 0.25 s of overlap or more than 1.0 s of response latency are excluded.

Each retained pair contains identifiers for the session, interaction, dyad, prompt speaker, response speaker, and turn index; timing fields for prompt and response boundaries; response latency and overlap; transcripts; relationship metadata; and the extracted metric fields.
The final pair identifier is a deterministic merge key built from session, interaction, dyad, speaker, turn, and rounded timing fields.
Identity-like columns are used for bookkeeping and split construction but are not used as direct predictive features in the evaluator, in order to reduce leakage from speaker or session identity.
The current pipeline uses a speaker-component split: dyads are represented as a graph over participant identifiers, connected components are assigned to train, validation, and test partitions with a fixed seed, and the target proportions are 60\%, 10\%, and 30\%.
This split prevents a connected speaker component from appearing in more than one partition.

\subsection{Prosodic and Temporal Metrics}
For each prompt-response segment, we compute prosodic measurements from the corresponding audio interval.
Fundamental frequency is estimated with Praat's autocorrelation pitch tracker through Parselmouth using a 75--500 Hz range \cite{jadoul2018introducing, vogel2009standardization}.
Only voiced frames are used for $F_0$ statistics.
The $F_0$ standard deviation and range are computed after trimming to the 10th--90th percentile interval, and voiced ratio records the fraction of voiced pitch frames.
Status fields mark segments with too little voiced support and can be excluded from model training, following the broader principle that unstable voice measurements should be surfaced rather than silently mixed into acoustic summaries \cite{titze_toward_1994}.

Temporal features are computed from word timings that fall inside the refined segment boundaries.
The local speech rate is the number of words per minute over the full segment duration, while articulation rate is divided by the non-pause duration.
Pauses are inter-word gaps of at least 0.2 s.
Pause ratio is total pause time divided by segment duration, and mean pause duration is the average duration of detected pauses, with zero used when no pause is present.
The 0.2 s threshold is a practical convention for silent pause analysis, and word-per-minute rate remains closely tied to perceived speech tempo \cite{campione2002large, iwarsson2023measuring}.
These computations are applied symmetrically to prompts and responses so that prompt-side timing can condition the model and response-side timing can be predicted and scored \cite{goldman1956determinants}.

\subsection{Vox-Profile-Derived Annotations}
The dataset does not provide all speaker and interaction-state annotations needed for conditional naturalness modeling.
We therefore augment the pair table with model-derived Vox-Profile annotations for age, gender posterior information, and dimensional affect \cite{feng2025vox}.
Audio is converted to mono, resampled to 16 kHz, RMS-normalized, and passed through Silero VAD to retain speech regions \cite{Silero_VAD}.
Speech-only audio is split into chunks of at most 15 s, retaining chunks of at least 3 s.

Static speaker-profile estimates are computed with a WavLM-based age and sex model over all speech chunks available for each speaker channel, then averaged to produce age, gender label, and gender confidence fields.
Dynamic affect estimates are computed for each prompt and response segment with a WavLM-based arousal-valence-dominance model, again averaging over valid speech chunks.
The age and gender outputs are treated as speech-model profile estimates rather than ground-truth demographic attributes.
Arousal, valence, and dominance are used as continuous descriptors of interactional state.
These conditioning variables are included because prior work links habitual $F_0$ to speaker traits, conversational timing to age and fluency-related behavior, and prosodic expressivity to emotional activation and dominance-related state \cite{traunmuller_acoustic_2000, tauroza_speech_1990, goldman1958speech, banziger_role_2005, banse1996acoustic, geng2020acoustic}.
Because these annotations are predicted rather than self-reported, final interpretation should distinguish model-conditioned stratification from claims about true demographic attributes.

\subsection{Merged Tabular Table}
The prosody/timing table and the Vox-Profile annotation table are merged with a strict one-to-one key over session, interaction, dyad, speaker roles, turn index, and rounded prompt and response boundaries.
The merge asserts that keys are unique in both sources and that shared bookkeeping columns match exactly.
Only pairs present in both tables are retained, and the global pair identifier is replaced with the deterministic merge key.
Status columns from each extraction stage are preserved so experiments can require valid prosody, timing, VAD, and Vox-Profile support.

\subsection{Current-Turn and History Features}
The current conditioning vector $x_t$ contains prompt-side speech measurements and available conversational context.
The hand-engineered prompt metrics include duration, word count, $F_0$ mean, $F_0$ standard deviation, $F_0$ range, voiced ratio, local speech rate, local articulation rate, pause ratio, mean pause duration, arousal, valence, dominance, estimated age, predicted gender information, gender confidence, Vox-Profile chunk count, and support speech duration.
Contextual fields include turn index, relationship metadata, and speaker-role fields.
The pair table also records latency and overlap at the prompt-response boundary for filtering, diagnostics, and evaluation settings where system turn timing is observed; strict prompt-only scoring excludes any response-derived transition quantity from $x_t$.
Numeric fields are standardized after imputation, and categorical fields are encoded for neural modeling.

The multiturn model augments $x_t$ with $h_t^{(H)}$, a padded sequence of up to $H$ previous turns from the same interaction.
History is grouped by interaction identifier and ordered by turn index.
For each current example, the history sequence contains the most recent previous turns, oldest to newest, with a Boolean padding mask for missing history slots.
A history turn contains the previous prompt-side tabular features, the embedding streams enabled for the current condition, and the previous response metric vector.

\subsection{Response Target Metrics}
The target vector $y_t \in \mathbb{R}^{15}$ contains response-side metrics chosen to describe conversational naturalness across lexical, prosodic, temporal, affective, and speaker-profile dimensions.
The 15 modeled metrics are response duration, word count, $F_0$ standard deviation, $F_0$ range, voiced ratio, speech rate, articulation rate, pause ratio, mean pause duration, arousal, valence, dominance, estimated age, gender confidence, and Vox-Profile support speech seconds.
The sweep excludes response $F_0$ mean and Vox-Profile chunk count from the target vector because they were degenerate or unstable targets in the current modeling setup.
The retained metrics match the dimensions used to score future S2S outputs, making the evaluator directly interpretable at the feature level.

\section{Modeling Approach}
\subsection{Problem Formulation}
For each prompt-response pair $t$, the model observes a current prompt-side conditioning vector $x_t$, a history tensor $h_t^{(H)}$, and a response-side target vector $y_t$.
The learning objective is to estimate the conditional distribution in \eqref{eq:conditional_distribution}.
For an S2S system output, the same response-side metrics are extracted to form $\hat{y}_{\mathrm{s2s},t}$, which is then compared against $p_{\theta}(y_t \mid x_t, h_t^{(H)})$ for the original prompt and history.

This setup separates generation from evaluation.
The model does not attempt to synthesize speech or text.
It learns what response metric distributions are plausible in natural human dialogue for a given prompt and recent conversational context.
The resulting evaluator can provide both a scalar plausibility score and per-metric diagnostics.

\subsection{Conditional Variational Autoencoder}
The CVAE introduces a latent variable $z$ to represent conditional response variability.
The encoder approximates $q_{\phi}(z \mid x_t,h_t^{(H)},y_t)$, and the decoder predicts a diagonal Gaussian distribution over $y_t$ from $(x_t,h_t^{(H)},z)$.
Training minimizes
\begin{equation}
    \mathcal{L} = 
    -\log p_{\theta}(y_t \mid x_t,h_t^{(H)},z)\\
    + \beta D_{\mathrm{KL}}\left(q_{\phi}(z \mid x_t,h_t^{(H)},y_t) \parallel p(z)\right),
    \label{eq:cvae_loss}
\end{equation}
where $p(z)$ is a standard normal prior and $\beta$ controls the strength of the KL regularization.
At evaluation time, samples from $p(z)$ yield Monte Carlo estimates of the predictive distribution.
This formulation is intended to capture one-to-many response structure: several response metric configurations may be natural for the same prompt and history.

\subsection{Multiturn History Encoder}
The reported experiments use a MultiTurnCVAE.
For each example, the padded history tensor is projected into a transformer history encoder with a learned CLS token.
The encoder uses two transformer layers, four attention heads, hidden size 256, GELU activations, dropout, and a padding mask that prevents missing history slots from contributing to the context representation.
If no previous turn is available, the history context is zeroed and the model falls back to current-turn conditioning.

The current-turn feature vector is projected separately, concatenated with the history context, and passed through a fusion multilayer perceptron to produce the final conditioning representation for the CVAE encoder and decoder.
The latent dimension is 16.
The decoder predicts both a mean and diagonal log variance for the 15 response targets, with log variance clamped for numerical stability.

\subsection{Implementation Details}
All variants use the same processed tabular conditioning and target matrices.
Numeric conditioning fields are median-imputed, augmented with missingness indicators where needed, and standardized.
Categorical relationship and predicted gender fields are one-hot encoded.
Targets are imputed and standardized for training, with original-scale metrics recovered for analysis when needed.

Training uses Adam with learning rate $10^{-3}$, weight decay $10^{-5}$, batch size 256, gradient clipping, and early stopping on validation loss.
The KL term is annealed from 0.001 to 0.05 during training.
The reported sweep evaluates the best checkpoint for each variant using 1,000 bootstrap samples to estimate 95\% confidence intervals.

\section{Embedding Conditioning}
\subsection{VoxProfile/Whisper Prompt Embeddings}
The embedding-enabled pipeline merges pair-level VoxProfile/Whisper representations with the tabular prompt features.
For each prompt segment, a 1,280-dimensional vector is loaded from the embedding table and concatenated with the processed conditioning representation.
These vectors are derived from hidden states of a Whisper-based speech model and are intended to encode broader information about the prompt signal than the hand-engineered metrics alone.
The embedding loader verifies one-to-one alignment with pair identifiers, checks that all vector files exist, and asserts a consistent embedding dimensionality before training.

\subsection{Trillsson Prompt Embeddings}
We also extract Trillsson4 embeddings for the paired prompt and response segments using the TensorFlow Hub Trillsson model.
For evaluator conditioning, the prompt-side Trillsson vector is used as an additional representation of the prompt.
The current extraction pipeline saves pooled 1,024-dimensional Trillsson vectors at 16 kHz for both prompt and response segments, with the prompt vector serving as a model input stream for $p_{\theta}(y_t \mid x_t,h_t^{(H)})$.
Trillsson is included because non-semantic speech representations are designed to preserve paralinguistic information such as speaker state, style, and acoustic context.

\subsection{Fusion Conditions}
The sweep compares four embedding conditions across all history windows.
The tabular condition uses only structured prompt and history features.
The Trillsson condition appends the 1,024-dimensional Trillsson vector to current and history turns.
The VoxProfile/Whisper condition appends the 1,280-dimensional VoxProfile/Whisper vector.
The dual-embedding condition appends both streams, allowing the model to use complementary ASR-derived and non-semantic speech representations.
All embedding streams are scaled before training.

\section{Scoring Framework}
\subsection{Likelihood and Residual Scores}
Given a prompt, its history, and an S2S response metric vector $\hat{y}_{\mathrm{s2s},t}$, probabilistic models can assign a negative log-likelihood score:
\begin{equation}
    \mathrm{NLL}(\hat{y}_{\mathrm{s2s},t},x_t,h_t^{(H)})
    =
    -\log p_{\theta}(\hat{y}_{\mathrm{s2s},t} \mid x_t,h_t^{(H)}).
    \label{eq:nll_score}
\end{equation}
Lower NLL indicates that the output lies in a higher-density region of the learned natural-response distribution.
For model families that output per-feature variance estimates, we also compute standardized residuals:
\begin{equation}
    r_j =
    \frac{\hat{y}_{\mathrm{s2s},t,j} - \mu_j(x_t,h_t^{(H)})}{\sigma_j(x_t,h_t^{(H)})}.
    \label{eq:standardized_residual}
\end{equation}
These residuals identify the direction and magnitude of each deviation.
For example, an output may be globally plausible but have an unusually long duration, compressed pitch range, or atypical pause ratio for its prompt and history.

\subsection{Interval and Calibration Diagnostics}
The evaluator should also report whether each observed metric lies inside model-predicted intervals.
For CVAE models, intervals can be estimated from Monte Carlo predictive samples.
Coverage at nominal levels such as 50\%, 80\%, 90\%, and 95\% provides a calibration diagnostic: a useful evaluator should not only predict accurate means, but also assign realistic uncertainty to natural response variability.

\subsection{Interpretable Aggregation}
A practical S2S diagnostic can aggregate likelihood, residual, and interval signals into a scalar naturalness score while preserving feature-level explanations.
The scalar score supports system comparison, while the per-feature diagnostics support development and error analysis.
This is important for S2S systems because unnaturalness is often localized: a response may have appropriate content and affect but fail through timing, pausing, or prosodic range.

\section{Experimental Protocol}
\subsection{Automatic Model Validation}
We evaluate MultiTurnCVAE variants across four history windows, $H \in \{1,3,5,7\}$, and four embedding conditions: tabular-only, Trillsson, VoxProfile/Whisper, and dual embedding.
Rows are grouped by interaction identifier and ordered by turn index to construct history sequences.
After status filtering and target exclusions, the sweep uses 39,127 training examples and 5,349 validation examples.
All variants use the same split policy, preprocessing state, target scaling, response-metric extraction procedure, architecture family, optimizer settings, and early-stopping rule.

We report scaled validation MSE, validation $R^2$, and C-index.
MSE and $R^2$ measure prediction accuracy in standardized target space.
C-index measures rank agreement and is included because a diagnostic evaluator may also be useful when it orders more natural responses above less natural ones.
The primary model-selection claim is based on MSE and $R^2$; C-index is analyzed separately because it does not follow the same ordering.

\subsection{Prompt-Response Human Evaluation}
We plan a listener study to test whether SPEAR scores track perceived contextual naturalness.
Each item presents a prompt audio clip followed by a candidate response, and raters score the response on 5-point MOS/ACR scales for overall naturalness, conversational fit to the prompt, timing and turn-taking, prosody, and signal quality.
The study will include 40 items from each of three conditions: natural human-human Seamless Interaction responses, deliberately unnatural EmotiVoice responses, and pooled human-to-S2S-agent responses.
Each item will receive at least four independent ratings.

The primary analysis will compare mean MOS with 95\% confidence intervals across conditions and correlate the automatic SPEAR score or NLL with the human ratings.
We will also test whether per-metric residuals explain listener comments about timing, pausing, speaking rate, prosody, or affect, and whether the automatic score remains informative after accounting for signal-quality ratings.
The human-to-S2S-agent condition is treated as exploratory rather than a ground-truth natural or unnatural class.

\section{Results}
\begin{table*}[t]
\centering
\caption{Multiturn CVAE validation sweep over embedding condition and history window. MSE is computed in standardized target space with 95\% bootstrap confidence intervals. The bold row is best by MSE and $R^2$.}
\label{tab:multiturn_sweep}
\begin{tabular}{llccc}
\toprule
Embedding & $H$ & MSE (95\% CI) & $R^2$ & C-index \\
\midrule
Tabular & 1 & 0.01513 [0.01441, 0.01586] & 0.98622 & 0.95982 \\
Tabular & 3 & 0.01311 [0.01257, 0.01367] & 0.98805 & 0.95800 \\
Tabular & 5 & 0.01245 [0.01186, 0.01306] & 0.98865 & 0.95454 \\
Tabular & 7 & 0.01322 [0.01258, 0.01390] & 0.98795 & 0.95541 \\
\midrule
Trillsson & 1 & 0.05256 [0.05093, 0.05422] & 0.95211 & 0.94311 \\
Trillsson & 3 & 0.01380 [0.01322, 0.01442] & 0.98742 & 0.95435 \\
Trillsson & 5 & 0.05329 [0.05117, 0.05569] & 0.95144 & 0.93721 \\
Trillsson & 7 & 0.05392 [0.05229, 0.05572] & 0.95087 & 0.93767 \\
\midrule
VoxProfile/Whisper & 1 & 0.01299 [0.01229, 0.01377] & 0.98816 & 0.95724 \\
VoxProfile/Whisper & 3 & 0.05450 [0.05293, 0.05614] & 0.95034 & 0.93681 \\
VoxProfile/Whisper & 5 & 0.01522 [0.01465, 0.01584] & 0.98614 & 0.95687 \\
VoxProfile/Whisper & 7 & 0.01674 [0.01601, 0.01755] & 0.98475 & 0.95388 \\
\midrule
Dual embed & 1 & 0.05379 [0.05204, 0.05556] & 0.95099 & 0.93954 \\
\textbf{Dual embed} & \textbf{3} & \textbf{0.01197 [0.01140, 0.01252]} & \textbf{0.98910} & 0.95749 \\
Dual embed & 5 & 0.01383 [0.01316, 0.01452] & 0.98740 & 0.95835 \\
Dual embed & 7 & 0.04921 [0.04727, 0.05124] & 0.95516 & 0.93812 \\
\bottomrule
\end{tabular}
\end{table*}

Table~\ref{tab:multiturn_sweep} shows that the best variant by MSE and $R^2$ is dual embedding with a three-turn history window.
This model achieves MSE 0.01197 and $R^2=0.98910$.

The strongest tabular-only baseline is the five-turn history model, with MSE 0.01245 and $R^2=0.98865$.
Thus, the dual-embedding $H=3$ model reduces MSE by 3.90\% relative to the strongest tabular-only multiturn model.
This is the only embedding condition and history window in the sweep that improves over the best tabular-only result by MSE.

The embedding results are strongly dependent on history length.
Trillsson is competitive at $H=3$ but degrades at $H=1$, $H=5$, and $H=7$.
VoxProfile/Whisper is strongest at $H=1$ and becomes weaker at longer windows.
Dual embedding performs best at $H=3$ but also degrades sharply at $H=1$ and $H=7$.
This pattern suggests that richer speech embeddings are useful when the history window is long enough to provide conversational context but not so long that the transformer history encoder must integrate noisy or weakly relevant prior turns.

C-index does not follow the same ordering as MSE and $R^2$.
The highest C-index is obtained by the tabular-only $H=1$ model, with C-index 0.95982.
Therefore, the dual-embedding $H=3$ model should be understood as the best predictive model by MSE and $R^2$, while tabular-only short-history conditioning gives the strongest rank-ordering result under C-index.


\subsection{Human Evaluation Results}
The human evaluation has not yet been run.  Table~\ref{tab:human_eval_placeholder} defines the planned reporting format for the prompt-response listener study.

\begin{table*}[t]
\centering
\caption{Planned human evaluation summary. Values will be reported as mean MOS with 95\% confidence intervals after annotation is complete.}
\label{tab:human_eval_placeholder}
\begin{tabular}{lccccc}
\toprule
Condition & Overall & Prompt fit & Timing & Prosody & Signal quality \\
\midrule
Natural human-human & -- & -- & -- & -- & -- \\
Neutral S2S-agent & -- & -- & -- & -- & -- \\
Unnatural EmotiVoice & -- & -- & -- & -- & -- \\
\bottomrule
\end{tabular}
\end{table*}

After ratings are collected, we will report condition-level MOS differences and correlations between human ratings and SPEAR scores.  The expected evidence pattern is that natural human-human responses receive the highest contextual naturalness ratings, EmotiVoice responses receive lower ratings, and S2S-agent responses occupy an exploratory middle region whose location depends on the system outputs.

\section{Conclusion}
This work develops a speech-native, prompt-conditioned framework for evaluating whether S2S responses fall inside natural conversational distributions.
We construct a paired prompt-response metric table from natural dyadic interactions, augment it with Vox-Profile-derived annotations and speech embeddings, and model response-side naturalness metrics with a multiturn CVAE.
The validation sweep shows that history length and embedding choice interact strongly: dual Trillsson plus VoxProfile/Whisper conditioning is beneficial at a three-turn window and is the only evaluated embedding setting that improves MSE over the strongest tabular-only model.
These results support multiturn, distributional evaluation as a practical route toward interpretable S2S naturalness scoring, while also showing that speech embeddings must be fused with an appropriate conversational context window to be useful.
The planned human evaluation will test whether these automatic scores align with listener judgments of contextual response naturalness and will provide the main subjective validation for future revisions.

\bibliographystyle{IEEEtran}
\bibliography{newbib}

\end{document}